---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
##Introduction
My gihub rep is here with all necessary code:
[Peer Assignment](https://github.com/brennap3/RepData_PeerAssessment1/blob/master/Roughwork.R)

This assignment also uses dplyr 0.5 (to make use of the coalesce function), which has just been released so if running the code please update your dplyr pacakge, the notes on new features are here:
[Dplyr new features](https://blog.rstudio.org/2016/06/27/dplyr-0-5-0/)

We use the following packages in oour analysis, so lets load them

```{r echo=TRUE, eval=TRUE, message=FALSE}
library(tibble)
library(dtplyr)
library(dplyr)
library(ggplot2)
library(magrittr)
```

## Loading and preprocessing the data
The data is loaded using the read.csv function and a summary is run of the data. From the summary we can see their are some problems with the data, there appears to be missing data. We can see some problems with our dataset, the dimension steps is missing 2304 values.

```{r echo=TRUE, eval=TRUE, message=TRUE}
datafitbit<-read.csv("C:\\Users\\Peter\\Desktop\\RepData_PeerAssessment1\\Data\\activity.csv")
summary(datafitbit)
```

## What is mean total number of steps taken per day?
To do this we will use dplyr we will calculate total number of steps per day, then from this calculate the mean is 10766.19, the median is 10765 and the standard deviation is 4269.18. 

```{r echo=TRUE, eval=TRUE, message=FALSE}
datastepsbydate<-datafitbit %>%
    group_by(date) %>%
      summarise( sum_steps = sum(steps)) %>%
        arrange(sum_steps) %>% as.data.frame()

avgdatastepsbydate<-mean(datastepsbydate$sum_steps,na.rm=TRUE) 
mediandatastepsbydate <-median(datastepsbydate$sum_steps,na.rm=TRUE) 
standard.deviation.datastepsbydate <-sd(datastepsbydate$sum_steps,na.rm=TRUE) 
print(avgdatastepsbydate)
print(mediandatastepsbydate)
print(standard.deviation.datastepsbydate)
```

Lets vizualize the distributions in base R and ggplot (using qplot) using a histogram.
```{r echo=TRUE, eval=TRUE, message=FALSE}
hist(datastepsbydate$sum_steps,main="Total number of steps \n taken per day",xlab="Sum of steps per day",ylab="Frequency")

qplot(datastepsbydate$sum_steps,
      geom="histogram",
      main = "Total number of steps \n taken per day",
      binwidth=5000,
      xlab="Sum of steps",
      ylab="Frequency"
      )

```

Lets also see what the distributions look like using a boxplot again in base r and ggplot (q plot), from the boxplots we can see the mean , 25th and 75th percentile (the IQR the inner quartile range) the whiskers extended to  Q1-1.5 X IQR and Q3+1.5XIQR, and there also seems to be a number (3) outliers. The median can be seen at the heavy black line.

```{r echo=TRUE, eval=TRUE, message=FALSE}

boxplot(datastepsbydate$sum_steps,main="Distribution of steps by Day",ylab="Sum of steps per day")
## or in ggplot using qqplot
ggplot(datastepsbydate, aes(factor(0),sum_steps))+geom_boxplot()+
  ggtitle("Total number of steps \n taken per day \n no missing data imputed")+
  xlab("")+
  ylab("Total number of steps taken per day")

```


## What is the average daily activity pattern?

We can look at the average daily activity by using the following procedure:
* Aggregate the data with dplyr by Interval and then calculate the mean
* Create a time series plot, using ggplot 

```{r echo=TRUE, eval=TRUE, message=FALSE}
##Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
##and the average number of steps taken, averaged across all days (y-axis)

datafitbit.avg.daily.activity<-datafitbit %>%
  group_by(interval) %>%
  summarise( mean_steps = mean(steps,na.rm=TRUE)) %>%
  arrange(interval) %>% as.data.frame()

ggplot(datafitbit.avg.daily.activity, aes(x=interval, y=mean_steps)) +
  geom_line(color = "blue")+
  ggtitle("time series plot of the 5-minute interval (x-axis) \n and the average number of steps taken, \n averaged across all days (y-axis)")+
  xlab("interval")+
  ylab("mean number of \n steps across all days")
```

Next we can identify the interval which has the highest mean number of steps, by selecting the row
from the aggregated dataset the highest average value by index.

```{r echo=TRUE, eval=TRUE, message=FALSE}
f0<-max(datafitbit.avg.daily.activity$mean_steps)

##where does the max average occur

datafitbit.avg.daily.activity[which(datafitbit.avg.daily.activity$mean_steps==eval(f0)),]
```


## Imputing missing values
Imputation of missing values was done according to the following procedure:
* Calculated the mean by interval
* That mean was steps grouped by interval was then merged to the original dataset
* Coalesce of steps with mean_steps_per_interval was then used to replace missing values in steps with the caclculated averages


```{r echo=TRUE, eval=TRUE, message=FALSE}
datafitbit.per.interval.activity<-datafitbit %>%
  group_by(interval) %>%
  summarise( mean_steps_per_interval = mean(steps,na.rm=TRUE)) %>%
  arrange(interval) %>% as.data.frame()

##now merge on data to original dataset

datafitbit.with.daily.averages<-dplyr::left_join(datafitbit,datafitbit.per.interval.activity,by="interval")

##now replace with mean value where blank

head(datafitbit.with.daily.averages)

##dplyr::coalesce(as.double(datafitbit.with.daily.averages$steps),datafitbit.with.daily.averages$mean_steps_per_interval)

##test works

##dplyr::coalesce(as.double(datafitbit.with.daily.averages$steps),datafitbit.with.daily.averages$mean_steps_per_interval)

datafitbit.with.daily.averages$steps<-dplyr::coalesce(as.double(datafitbit.with.daily.averages$steps),datafitbit.with.daily.averages$mean_steps_per_interval)

##test works

head(datafitbit.with.daily.averages$steps)
summary(datafitbit.with.daily.averages$steps)
```
On imputing the missing data we visualize the data by again creating a histogram in base r and ggplot.

```{r echo=TRUE, eval=TRUE, message=FALSE}
#Make a histogram of the total number of steps taken each day 
head(datafitbit.with.daily.averages)


datastepswithimputedmissingdatebydate <- datafitbit.with.daily.averages %>%
  group_by(date) %>%
    summarise( sum_steps = sum(steps)) %>%
        arrange(sum_steps) %>% as.data.frame()


hist(datastepswithimputedmissingdatebydate$sum_steps, main="Total number of steps \n taken per day \n missing values imputed", xlab="Sum of steps per day",ylab="Frequency")
```
The histogram in ggplot.

```{r echo=TRUE, eval=TRUE, message=FALSE}

qplot(datastepswithimputedmissingdatebydate$sum_steps,
      geom="histogram",
      main = "Total number of steps \n taken per day \n with missing data imputed",
      binwidth=5000,
      xlab="Sum of steps",
      ylab="Frequency"
)

```

##Calculate and report the mean and median total number of steps taken per day. 

The mean and median total number of steps  taken per day based on the dataset with the imputed data replacing the missing data are shown below. 


```{r echo=TRUE, eval=TRUE, message=FALSE}
avg.datasteps.bydate.withimputedmissingdata<-mean(datastepswithimputedmissingdatebydate$sum_steps,na.rm=TRUE) 
median.datasteps.bydate.withimputedmissingdata<-median(datastepswithimputedmissingdatebydate$sum_steps,na.rm=TRUE) 
sd.datasteps.bydate.withimputedmissingdata<-sd(datastepswithimputedmissingdatebydate$sum_steps,na.rm=TRUE)

print(avg.datasteps.bydate.withimputedmissingdata)
print(median.datasteps.bydate.withimputedmissingdata)
print(sd.datasteps.bydate.withimputedmissingdata)

```

##Do these values differ from the estimates from the first part of the assignment? 

To answer the question do these values differ we create a number of side by side plotsshowing the distibutions of (a) The dataset without the missing data and (b) with the missing data. We use two types of distribution plots to do this;

1. Histograms.
2. Boxplots.

```{r  echo=TRUE, eval=TRUE, message=FALSE}
##Do these values differ from the estimates from the first part of the assignment? 
library(cowplot)

ap<-qplot(datastepsbydate$sum_steps,
          geom="histogram",
          main = "Total number of steps \n taken per day",
          binwidth=5000,
          xlab="Sum of steps",
          ylab="Frequency"
)

bp<-qplot(datastepswithimputedmissingdatebydate$sum_steps,
          geom="histogram",
          main = "Total number of steps \n taken per day \n with missing data imputed",
          binwidth=5000,
          xlab="Sum of steps",
          ylab="Frequency"
)
  
plot_grid(ap, bp, ncol = 2, nrow = 1)

```

The boxplots.

```{r  echo=TRUE, eval=TRUE, message=FALSE}

abxpwithmiss<-ggplot(datastepswithimputedmissingdatebydate, aes(factor(0),sum_steps))+geom_boxplot()+
  ggtitle("Total number of steps \n taken per day \n with missing data imputed")+
    xlab("")+
      ylab("Total number of steps taken per day")
  
abxpw<-ggplot(datastepsbydate, aes(factor(0),sum_steps))+geom_boxplot()+
  ggtitle("Total number of steps \n taken per day \n no missing data imputed")+
  xlab("")+
  ylab("Total number of steps taken per day")


plot_grid(abxpwithmiss,abxpw,  ncol = 2, nrow = 1)

```

The summary statistics are shown below:

Summary statistics.

```{r  echo=TRUE, eval=TRUE, message=FALSE}
summary(datastepsbydate)
summary(datastepswithimputedmissingdatebydate)
```

We can see from examining the distributions (with both boxplots and histograms) and looking at the summary statistics that with the imputed missing data, that the IQR has increased, but there does appear to beslightly more left skew in our data and also our distribution with the imputed data appears to be more peaked. We will calculate skew and also the kurtosis the peakedness of the distribution.

```{r  echo=TRUE, eval=TRUE, message=FALSE}
summary(datastepsbydate)
summary(datastepswithimputedmissingdatebydate)
```

The skewness and the kurtosis of the original dataset and the dateaset with the imputed missing data are calculated below: 

With the Missing Data:
The skewness value caculated is negative  (-0.332957), indicating the distribution is skewed left which we could observe from distribution plots, we can also see from the kurtosis values that distribution is somewhat more peaked (when compared to a normal distribution)

```{r  echo=TRUE, eval=TRUE, message=FALSE}
library(moments)
skewness(datastepsbydate$sum_steps,na.rm=T)
kurtosis(datastepsbydate$sum_steps,na.rm=T)
```

When we again measure the skewness and kurtosis of the dataset, we slightly higher negative skew (skewed left), but the Kurtosis value is much higher indicating that the distribution is more peaked. Again on examining the previously calculated distribution plots we can confirm this observation.


```{r  echo=TRUE, eval=TRUE, message=FALSE}
library(moments)
skewness(datastepswithimputedmissingdatebydate$sum_steps,na.rm=T) ##no need to do this they (NA's) should be removed
kurtosis(datastepswithimputedmissingdatebydate$sum_steps,na.rm=T) ##no need to do this they  (NA's)  should be removed
```

With the imputed missing Data:
## Are there differences in activity patterns between weekdays and weekends?

To calculate whether there are differences in behaviour we carry out the following procedure:
* Calculate the weekday from the date by using lubridate
* Calculate mean number of step by day of week and interval 
* Create a ggplot heatmap visualizing the number of teps taken per day.

From this analysis its quite clear that peak activity during oocurs between 8-9 am on weekdays (Monday to Friday) and that on weekends peak activity  occurs much later in the day. The code to execute the analysis is shown below.

```{r  echo=TRUE, eval=TRUE, message=FALSE}
library(lubridate)
library(RColorBrewer)
head(datafitbit)
datafitbit['Week_Day']<-lubridate::wday(datafitbit$date,label=T)

## we will use the dataset with out the imputed data

datafitbit.per.interval.activity.by.weekday<-datafitbit %>%
  group_by(interval,Week_Day) %>%
  summarise( mean_steps_per_interval = mean(steps,na.rm=TRUE)) %>%
  arrange(Week_Day,interval) %>% as.data.frame()

##head(datafitbit.per.interval.activity.by.weekday)


jBuPuFun <- colorRampPalette(brewer.pal(n = 9, "BuPu"))
paletteSize <- 256
jBuPuPalette <- jBuPuFun(paletteSize)

ggplot(datafitbit.per.interval.activity.by.weekday, aes(x = interval, y = Week_Day, fill = mean_steps_per_interval)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_tile() +
  ggtitle("Heatmap of mean number of steps \n by time and day of Week")+
  xlab("Time of day \n 24 hour clock")+
  ylab("Day") +
  scale_fill_gradient2(low = jBuPuPalette[1],
                       mid = jBuPuPalette[paletteSize/2],
                       high = jBuPuPalette[paletteSize],
                       midpoint = (max(datafitbit.per.interval.activity.by.weekday$mean_steps_per_interval) + min(datafitbit.per.interval.activity.by.weekday$mean_steps_per_interval)) / 2,
                       name = "Mean steps per interval \n per day of week")
```

Again if we code the days of week as weekend (Saturday, Sunday) and again plot the x -axis (Monday, Tuesday,Wednesday,Thursday,Friday) as time of day, the pattern is somewhat clearer, on weekdays the most activity is happening between 8 am and 9 am while on weekends activity is occuring much later.

```{r  echo=TRUE, eval=TRUE, message=FALSE}

datafitbit$Week_day_Week_End <- sapply(datafitbit$Week_Day, function(x) switch(as.character(x),
                                                             "Sun" = "Weekend",
                                                             "Sat" = "Weekend",
                                                             "Weekday"
))

datafitbit.per.interval.activity.by.weekday.weekend<-datafitbit %>%
  group_by(interval,Week_day_Week_End) %>%
  summarise( mean_steps_per_interval = mean(steps,na.rm=TRUE)) %>%
  arrange(Week_day_Week_End,interval) %>% as.data.frame()


ggplot(datafitbit.per.interval.activity.by.weekday.weekend, aes(x = interval, y = Week_day_Week_End, fill = mean_steps_per_interval)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
  geom_tile() +
  ggtitle("Heatmap of mean number of steps \n by time and day of Week type")+
  xlab("Time of day \n 24 hour clock")+
  ylab("Day") +
  scale_fill_gradient2(low = jBuPuPalette[1],
                       mid = jBuPuPalette[paletteSize/2],
                       high = jBuPuPalette[paletteSize],
                       midpoint = (max(datafitbit.per.interval.activity.by.weekday$mean_steps_per_interval) + min(datafitbit.per.interval.activity.by.weekday$mean_steps_per_interval)) / 2,
                       name = "Mean steps per interval \n per day of week type")
```

